Structuring College Access: The Market Segment Model and College Board Geomarkets

### 2025-07-22 text/manuscript-full-KS.qmd
blind authors

Geomarkets carve states and metroplitan areas into smaller geographic units. Upon closer inspection, the three orders selected on all Geomarkets in the Chicago-land area except for "IL10 - City of Chicago", which is significantly poorer and less White. In the DC/Maryland/Virginia metro area, the orders selected the affluent "MD02 - Montgomery Metropolitan," "MD03 - Central Maryland (w/o Baltimore)," "VA01 - Arlington & Alexandria," and "VA02 - Fairfax County" Geomarkets, but excluded the "MD05 - Prince Georges Metropolitan," "MD07 - Baltimore (Urban)," and "DC01 - District of Columbia" Geomarkets, which are poorer and have a higher share of Black residents. An analysis of 500 \[check\] student lists purchased by 9 public research universities found that 11% of purchases filtered on Geomarket [@RN5012]. Recent research suggests that student lists are surprisingly important for college access outcomes and college graduation outcomes, particularly for first-generation and underrepresented minority students [@RN4739; @RN4894]. To the extent that Geomarket borders are correlated with race and class, the Geomarket selections observed here raise concern that Geomarkets -- and other geographic search filters -- may systematically filter working class and non-white students from educational opportunities.
-->
```
Sociology offers several perspectives to describe the process of sorting students into colleges as a process that tends toward social reproduction [@RN4873; @RN2377]. On the student demand side, the status attainment model argues that student postsecondary destination is a function of family socioeconomic background, particularly parental education [e.g., @RN915; @RN1273; @RN1268; @RN5033; @RN1300]. The cultural capital model argues that bourgeois households bestow the pedigree, social networks, and information necessary to claim spots at selective colleges [e.g., @RN1228; @RN1241; @RN4991]. <!--@RN4873 describes schools as "sorting machines" that create, incorporate, and assign students to categories that determine educational opportunities, often reinforcing reinforcing racial and class inequality. -->On the supply-side, the credentialism literature recognizes that colleges have a financial incentive for educational credentials to be required to claim advantageous social and economic positions [@RN1297], and this race for credentials benefits affluent households [e.g., @RN865; @RN861]. Scholarship on enrollment management offers an agentic depiction of college behavior, showing that colleges expend considerable resources recruiting desirable prospects [@RN3519; @RN4758; @cottom2017lower; @RN4324].

@@

Third, Geomarkets were incorporated into College Board's student list product, Student Search Service. Student lists contain the contact information of prospective students and have been the primary source of "lead generation" in U.S. higher education since 1972, when College Board began selling names [@belkin2019-studata; @list_biz; @RN4728]. <!-- @RN4895 reported that 87% of private and 86% of public four-year institutions purchase student lists. --> Colleges control which prospect profiles they purchase by selecting search filters, such as SAT score, AP score, and state. Geomarket filters enable colleges to include or exclude prospects from particular Geomarkets.

@@

Scholarship on enrollment management has focused on the behavior of colleges [e.g., @cottom2017lower; @RN4758; @RN4759; @RN3519], contributing to the belief that recruiting is a function of individual colleges. However, scholarship has failed to investigate the broader enrollment management industry -- consisting of third-party vendors, their products, their consulting solutions -- as a set of mechanisms that structure college access. These mechanisms, which lie upstream of the behavior of individual colleges, are incorporated by individual colleges in ways that funnel certain kinds of students to certain kinds of institutions. This omission is surprising because sociology is concerned with social sorting [@RN4873], particularly sorting into college.

@@

Our analyses address two research questions, which speak to how Geomarkets are utilized within EPS software (RQ1) and within the Student Search Service student list product (RQ2). First, what is the socioeconomic and racial variation between Geomarkets in metropolitan areas and how does this variation change over time? We address this question by spatially joining Geomarket shapefiles to Census data about socioeconomic and racial characteristics. Second, how does the socioeconomic and racial composition of included versus excluded prospects vary when student list purchases filter on particular Geomarkets? This research question is motivated by @RN4982, which recommend that selective colleges target affluent Geomarkets, and by prior research on enrollment management behavior, which finds that selective colleges focus recruiting visits on high schools in affluent, predominantly White communities [@RN3519; @RN4758]. We analyze this research question using data from actual student list purchases, which we collected using public records requests. We analyze student list purchases that include all Geomarkets in a metropolitan area in order to investigate which prospects would be included or excluded had the student list purchase filtered on particular Geomarkets.

@@

@fig-em-funnel depicts the "enrollment funnel," which modifies the traditional marketing funnel to depict broad stages in the process of recruiting students [@RN4986; @RN2004]. The funnel begins with a large pool of "prospects" (i.e., prospective students) that the university would like to enroll. "Leads" are prospects whose contact information has been obtained. "Inquiries" are prospects that contact the institution, including those that respond to an initial solicitation (e.g., email) and those that reach out on their own (e.g., sending SAT scores). <!--  Applicants consist of inquiries who apply plus "stealth applicants" who do not contact the university before applying. The funnel narrows at each successive stage in order to convey the assumption of "melt" (e.g., a subset of "inquiries" will apply).  -->The purpose of the enrollment funnel is to inform recruiting interventions that target one or more stages. These interventions seek to increase the probability of "conversion" across stages [@RN4322]. At the top of the enrollment funnel, purchasing student lists is the primary means of converting prospects to leads [@list_biz]. Purchased leads are served emails, brochures, and targeted social media designed to solicit inquiries and applications [@RN4895].

@@

A growing literature analyzes the earlier "recruiting" stages of identifying leads, soliciting inquiries, and soliciting applications. @RN4758 conceptualize recruiting behavior as an indicator of college enrollment priorities. Ethnographies by @RN3519 and @RN4407 identify connections between private school guidance counselors and college admissions officers as a mechanism for social reproduction. In turn, recruiting visits to high schools -- and subsequent phone calls -- are a means of maintaining ties with guidance counselors at feeder schools and establishing relationships with prospective students [@RN3519; @RN4895]. Quantitative case-studies of off-campus recruiting visits by public research universities and by selective private universities reveal a preference for visiting private schools and affluent, predominantly White public schools [@RN4758; @RN4759; @RN5023]. From the student perspective, @RN4324 finds that underrepresented students were drawn to colleges that made them feel wanted, often attending institutions with lower graduation rates and requiring larger loans than other college options. @cottom2017lower shows that for-profit colleges found a niche in Black and Latinx communities because traditional colleges ignored these communities. <!--  Consistent with scholarship on political economy [@RN4517; @cottom2017lower; @RN5015], this literature reveals a vertical hierarchy of postsecondary institutions matching to a vertical hierarchy of customer niches segmented by class and race, a process consistent with the notion of that higher education is mechanism of social reproduction [@RN2377; @RN4873].   -->

@@

Drawing from scholarship on organizational theory, EM processes require colleges to make many "make or buy" [@RN4738; @RN980] decisions about whether to perform a given task in-house or outsource it to a third-party vendor [@list_biz]. The most widely-known class of EM vendor is EM consulting firms, which provide advice and implementation in the areas of marketing, recruiting, pricing and financial aid, and student success [@RN5027]. As "creating a class" has become complicated and high-stakes [@RN3519], many colleges hire EM consulting firms to develop and/or implement recruiting campaigns. The two largest firms -- Ruffalo Noel Levitz and EAB -- claim to serve more than 3,000 colleges and universities collectively [@rnlabout; @eabcomm].[^2]

[^2]: In our data collection, attempting to obtain data about student list purchases from all public universities in four states, at least half of these universities outsourced student list purchases to an EM consultancy [@list_biz; @list_empirics].

While the contemporary EM industry is characterized by software-as-a-service products sold to colleges by private equity backed firms [@list_biz], College Board products played a pivotal role in transforming recruiting from an in-house process to a process structured by third-party products. Every year, colleges must figure out which prospects should receive promotional material and which high schools to visit. The College Board Student Search Service -- copied by ACT's Educational Opportunity Service -- became the ubiquitous means of converting prospects into leads that could be served targeted recruiting interventions. The College Board Enrollment Planning Service (EPS), launched in 1984, is an early software-as-a-service product that provides recommendations about which Geomarkets colleges should recruit from and which high schools they should visit within targeted Geomarkets. Both products are applications of quantification that sort students into colleges. Student Search Service includes or excludes prospects from student list purchases by colleges based on the value of selected filters like Geomarket while EPS encourages colleges to target Geomarkets and high schools based on the characteristics of households. The multi-disciplinary literature on quantification helps conceptualize how third-party products utilize Geomarkets and the potential consequences of this utilization on the inequitable sorting of students into colleges.

@@

**Geomarket Filter in Student Search Service**. <!-- By 1984, College Board Geomarkets were incorporated as a search filter in the College Board Student Search Service \[CITE\]. --> Geomarkets also function as a student list search filter for the College Board's Student Search Service. A student list contains the contact information of prospective students who meet the search filter criteria (e.g., test score, GPA) specified by the university. Student lists are the fundamental input for undergraduate recruiting campaigns because purchased names -- alongside prospects who reach out on their own -- constitute the set of prospects who receive subsequent recruiting interventions (e.g., mail, email) designed to push them toward the application and enrollment stages of the "enrollment funnel." @RN4895 reports that 86% of public colleges and 87% of private colleges purchase student lists.[^11] Historically, dominant list vendors are the College Board and ACT, which derived student list data from test-takers, but the test-optional movement, advances in technology, and surging private equity investments have contributed to new sources of student list data [@list_biz].

@@

However, @RN5012 argue that student list products may exacerbate racial inequality in college access. The authors conceptualize student list products as "selection devices" [@RN4778] that enable colleges to select which prospective students they target by incorporating search filters (e.g., high school graduating class, state). @RN4786 defines "racialized inputs" as ostensibly race-neutral inputs that are systematically correlated with race because marginalized racial/ethnic groups have historically been excluded from the input. @RN5012 argue that several frequently utilized student list filters (e.g., zip code, AP test score, SAT score) meet the criteria of racialized inputs. Using a national sample of high school students and using data from actual student lists purchased by public universities, @RN5012 show that racialized search filters have a strong negative relationship with the selection of Black and Latinx prospects.

Because of the extent of residential segregation in the U.S., geographic borders are a commonly studied racialized input in scholarship about algorithmic bias [@RN4775; @RN4849; @RN4887]. Student list products offer many geographic filters. Some geographic filters are based on known geographic borders (e.g., zip code, county, CBSA, state). College Board also created geographic borders that subsequently became filters in the Student Search Service, such as Geomarkets and Geodemographic Segment filters.[^14] In their sample, @RN5012 found that 11% of student lists purchased by public research universities filtered on Geomarkets. Drawing from @RN4786, we argue that geomarkets satisfy the two criteria of racialized inputs. First, they are ostensibly race-neutral inputs in that neither @RN4982 -- nor subsequent EPS promotional material [e.g., @RN1621] -- mention race. Second, Geomarkets may be systematically correlated with race. For example, Geomarket "CA07" is "City of Oakland" and "CA08" is "Alameda County, excluding Oakland." <!--# A university might purchase prospect contact information by filtering on CA08 in combination with additional filters, such as year of high school graduation and SAT test score.--> We extend @RN5012 by examining how the racial composition of actual student list purchases changes if universities filter on particular Geomarkets. <!--# This manuscript focuses on Geomarkets. -->

@@

The discussion of homophily by @RN4975 describes not only the logic of the Market Segment Model but also a broader institutional logic of the enrollment management industry. The logic of homophily assumes that there is a natural cultural match between students and colleges, suggesting colleges should find prospective students who resemble students who enrolled at their institution in the past and focus recruiting efforts on the same schools and communities that are popular with peer colleges. We suggest that homophily is the operating logic most consistent with college recruiting behavior observed in the wild. For example, @RN5023 find great overlap in the sets of private schools visited by the University of South Carolina and the University of Alabama. <!--# The study also finds great overlap in the schools visited by Villanova University and the University of Notre Dame, with a disproportionate focus on visits to Catholic schools. When we observe that students at selective private colleges tend to be rich and that out-of-state students at public research universities tend to be rich [@RN4722], this is partly because these universities devote most of their recruiting expenditures towards students from affluent schools and communities [@RN4758; @RN5023]. --> As @RN4975 argues, observed homophily is not the result of voluntary action; rather, homophily is programmed into algorithms that create connections.

@@

Supply-side explanations of credentialism and more recent scholarship on enrollment management complement the cultural capital model of sorting students to colleges. The credentialism literature recognizes that colleges have a financial incentive for educational credentials to determine the competition for socially and economically advantageous positions [@RN865; @RN861; @RN880; @RN869] and this credentials arms race favors affluent households. Scholarship on enrollment management describes which students colleges want and how colleges go about attracting these students at different stages of the enrollment funnel [@RN3519; @RN4324; @RN3522; @RN4407; @RN4758; @RN3544; @RN3547; @cottom2017lower]. @RN3519 and @RN4321 describe a tacit arrangement between high school guidance counselors on the demand side and college admissions officers on the supply side. Counselors at well-resourced -- especially private -- schools are motivated to give their students a competitive advantage in admissions. Meanwhile, college admissions officers are motivated to enroll students who can afford full tuition price and are likely to donate in the future. These mutually beneficial desires are consummated by recognizing that upper and upper-middle class applicants satisfy the extracurricular needs of the college. The orchestra needs oboists. The lacrosse team needs players. Quantitative analyses of visits from colleges to high schools are consistent with @RN3519; selective private colleges and public research universities devote most of their recruiting resources on courting students from privileged schools and communities [@RN5023; @RN4758]. However, these analyses conceive of recruiting as something that is done solely by individual colleges.

@@

@list_biz describes four key dynamics in the enrollment management (EM) industry, with a focus on the market for student list data. First, EM consulting firms are central to the creation and implementation of recruiting campaigns [@RN5027]. The top two EM consultancies are Ruffalo Noel Levitz, which claims to serve 1,900 colleges annually [@rnlabout_2025], and EAB, which claims to serve 2,100 colleges [@about_eab]. College reliance on EM consultancies is partially explained by the high level of burnout and employee turnover in the EM profession [@hoover_em]. When we issued records requests to public universities about their student list purchases, at least 50% of universities indicated that they outsourced student list purchases to an EM consultancy. Often, the university was buying names each year but no employee had knowledge about which vendors they were buying from or which search filters they were utilizing.

@@

The third dynamic in the EM industry has been the growth of the test-optional admissions movement, which reduces both the number of paying test-takers and the coverage of student list products sold by College Board and ACT. ACT responded to uncertainty in their core testing business by attempting to become an edtech company [@blumenstyk_2019], failing [@molnar_2020a], and being acquired by Nexus Capital in 2024 [@knox_2024]. Over the last two decades, the testing companies developed new search filters that explicitly incorporated predictive analytics -- for example College Board's Geodemographic Segment filters [@RN4565] -- developed free college search engines, and entered the enrollment management consulting market more aggressively [@list_biz]. College Board has weathered the test-optional movement thanks to robust revenues from the Advanced Placement product, but the decline in PSAT/SAT/PreACT/ACT test-takers has undermined their oligopoly in the market for student list data.

The fourth -- and most important -- dynamic is the transformation of EM from an owner-operated to private-equity owned industry and the resulting eroding distinction between consultancy and software vendor. The 1980s and 1990s was a period of market entry in EM, when college admissions professionals or professors hung up a shingle [@RN5027]. The 2010s were a period of acquisitions and concentration. Drawing from resource dependence theory, @list_biz describe the proliferation of horizontal acquisitions to grow market share and vertical acquisitions designed to make customers more dependent on particular recruiting products. After regulatory scrutiny pushed private-equity interests out of the for-profit college market [@RN5022], private equity investors found value in acquiring firms that provided services to direct providers. Direct providers are subsidized by state and federal funds (Title IV financial aid) and these funds are then paid to third-party vendors. Therefore, private equity ownership of direct providers and third-party vendors are both examples of government-subsidized rent-seeking.

>>>>

Geomarkets carve states and metroplitan areas into smaller geographic units. Upon closer inspection, the three orders selected on all Geomarkets in the Chicago-land area except for "IL10 - City of Chicago", which is significantly poorer and less White. In the DC/Maryland/Virginia metro area, the orders selected the affluent "MD02 - Montgomery Metropolitan," "MD03 - Central Maryland (w/o Baltimore)," "VA01 - Arlington & Alexandria," and "VA02 - Fairfax County" Geomarkets, but excluded the "MD05 - Prince Georges Metropolitan," "MD07 - Baltimore (Urban)," and "DC01 - District of Columbia" Geomarkets, which are poorer and have a higher share of Black residents. An analysis of 500 \[check\] student lists purchased by 9 public research universities found that 11% of purchases filtered on Geomarket [@RN5012]. Recent research suggests that student lists are surprisingly important for college access outcomes and college graduation outcomes, particularly for first-generation and underrepresented minority students [@RN4739; @RN4894]. To the extent that Geomarket borders are correlated with race and class, the Geomarket selections observed here raise concern that Geomarkets -- and other geographic search filters -- may systematically filter working class and non-white students from educational opportunities.
-->
```
Sociology offers several perspectives to describe the process of sorting students into colleges as a process that tends toward social reproduction [@RN4873; @RN2377]. On the student demand side, the status attainment model argues that student postsecondary destination is a function of family socioeconomic background, particularly parental education [e.g., @RN915; @RN1273; @RN1268; @RN5033; @RN1300]. The cultural capital model argues that bourgeois households bestow the pedigree, social networks, and information necessary to claim spots at selective colleges [e.g., @RN1228; @RN1241; @RN4991]. <!--@RN4873 describes schools as "sorting machines" that create, incorporate, and assign students to categories that determine educational opportunities, often reinforcing reinforcing racial and class inequality. -->On the supply-side, the credentialism literature recognizes that colleges have a financial incentive for educational credentials to be required to claim advantageous social and economic positions [@RN1297], and this race for credentials benefits affluent households [e.g., @RN865; @RN861]. Scholarship on enrollment management offers an agentic depiction of college behavior, showing that colleges expend considerable resources recruiting desirable prospects [@RN3519; @RN4758_blinded; @cottom2017lower; @RN4324].

@@

Third, Geomarkets were incorporated into College Board's student list product, Student Search Service. Student lists contain the contact information of prospective students and have been the primary source of "lead generation" in U.S. higher education since 1972, when College Board began selling names [@belkin2019-studata; @list_biz_blinded; @RN4728]. <!-- @RN4895 reported that 87% of private and 86% of public four-year institutions purchase student lists. --> Colleges control which prospect profiles they purchase by selecting search filters, such as SAT score, AP score, and state. Geomarket filters enable colleges to include or exclude prospects from particular Geomarkets.

@@

Scholarship on enrollment management has focused on the behavior of colleges [e.g., @cottom2017lower; @RN4758_blinded; @RN4759_blinded; @RN3519], contributing to the belief that recruiting is a function of individual colleges. However, scholarship has failed to investigate the broader enrollment management industry -- consisting of third-party vendors, their products, their consulting solutions -- as a set of mechanisms that structure college access. These mechanisms, which lie upstream of the behavior of individual colleges, are incorporated by individual colleges in ways that funnel certain kinds of students to certain kinds of institutions. This omission is surprising because sociology is concerned with social sorting [@RN4873], particularly sorting into college.

@@

Our analyses address two research questions, which speak to how Geomarkets are utilized within EPS software (RQ1) and within the Student Search Service student list product (RQ2). First, what is the socioeconomic and racial variation between Geomarkets in metropolitan areas and how does this variation change over time? We address this question by spatially joining Geomarket shapefiles to Census data about socioeconomic and racial characteristics. Second, how does the socioeconomic and racial composition of included versus excluded prospects vary when student list purchases filter on particular Geomarkets? This research question is motivated by @RN4982, which recommend that selective colleges target affluent Geomarkets, and by prior research on enrollment management behavior, which finds that selective colleges focus recruiting visits on high schools in affluent, predominantly White communities [@RN3519; @RN4758_blinded]. We analyze this research question using data from actual student list purchases, which we collected using public records requests. We analyze student list purchases that include all Geomarkets in a metropolitan area in order to investigate which prospects would be included or excluded had the student list purchase filtered on particular Geomarkets.

@@

@fig-em-funnel depicts the "enrollment funnel," which modifies the traditional marketing funnel to depict broad stages in the process of recruiting students [@RN4986; @RN2004]. The funnel begins with a large pool of "prospects" (i.e., prospective students) that the university would like to enroll. "Leads" are prospects whose contact information has been obtained. "Inquiries" are prospects that contact the institution, including those that respond to an initial solicitation (e.g., email) and those that reach out on their own (e.g., sending SAT scores). <!--  Applicants consist of inquiries who apply plus "stealth applicants" who do not contact the university before applying. The funnel narrows at each successive stage in order to convey the assumption of "melt" (e.g., a subset of "inquiries" will apply).  -->The purpose of the enrollment funnel is to inform recruiting interventions that target one or more stages. These interventions seek to increase the probability of "conversion" across stages [@RN4322]. At the top of the enrollment funnel, purchasing student lists is the primary means of converting prospects to leads [@list_biz_blinded]. Purchased leads are served emails, brochures, and targeted social media designed to solicit inquiries and applications [@RN4895].

@@

A growing literature analyzes the earlier "recruiting" stages of identifying leads, soliciting inquiries, and soliciting applications. @RN4758_blinded conceptualize recruiting behavior as an indicator of college enrollment priorities. Ethnographies by @RN3519 and @RN4407 identify connections between private school guidance counselors and college admissions officers as a mechanism for social reproduction. In turn, recruiting visits to high schools -- and subsequent phone calls -- are a means of maintaining ties with guidance counselors at feeder schools and establishing relationships with prospective students [@RN3519; @RN4895]. Quantitative case-studies of off-campus recruiting visits by public research universities and by selective private universities reveal a preference for visiting private schools and affluent, predominantly White public schools [@RN4758_blinded; @RN4759_blinded; @RN5023_blinded]. From the student perspective, @RN4324 finds that underrepresented students were drawn to colleges that made them feel wanted, often attending institutions with lower graduation rates and requiring larger loans than other college options. @cottom2017lower shows that for-profit colleges found a niche in Black and Latinx communities because traditional colleges ignored these communities. <!--  Consistent with scholarship on political economy [@RN4517; @cottom2017lower; @RN5015], this literature reveals a vertical hierarchy of postsecondary institutions matching to a vertical hierarchy of customer niches segmented by class and race, a process consistent with the notion of that higher education is mechanism of social reproduction [@RN2377; @RN4873].   -->

@@

Drawing from scholarship on organizational theory, EM processes require colleges to make many "make or buy" [@RN4738; @RN980] decisions about whether to perform a given task in-house or outsource it to a third-party vendor [@list_biz_blinded]. The most widely-known class of EM vendor is EM consulting firms, which provide advice and implementation in the areas of marketing, recruiting, pricing and financial aid, and student success [@RN5027]. As "creating a class" has become complicated and high-stakes [@RN3519], many colleges hire EM consulting firms to develop and/or implement recruiting campaigns. The two largest firms -- Ruffalo Noel Levitz and EAB -- claim to serve more than 3,000 colleges and universities collectively [@rnlabout; @eabcomm].[^2]

[^2]: In our data collection, attempting to obtain data about student list purchases from all public universities in four states, at least half of these universities outsourced student list purchases to an EM consultancy [@list_biz_blinded; @list_empirics_blinded].

While the contemporary EM industry is characterized by software-as-a-service products sold to colleges by private equity backed firms [@list_biz_blinded], College Board products played a pivotal role in transforming recruiting from an in-house process to a process structured by third-party products. Every year, colleges must figure out which prospects should receive promotional material and which high schools to visit. The College Board Student Search Service -- copied by ACT's Educational Opportunity Service -- became the ubiquitous means of converting prospects into leads that could be served targeted recruiting interventions. The College Board Enrollment Planning Service (EPS), launched in 1984, is an early software-as-a-service product that provides recommendations about which Geomarkets colleges should recruit from and which high schools they should visit within targeted Geomarkets. Both products are applications of quantification that sort students into colleges. Student Search Service includes or excludes prospects from student list purchases by colleges based on the value of selected filters like Geomarket while EPS encourages colleges to target Geomarkets and high schools based on the characteristics of households. The multi-disciplinary literature on quantification helps conceptualize how third-party products utilize Geomarkets and the potential consequences of this utilization on the inequitable sorting of students into colleges.

@@

**Geomarket Filter in Student Search Service**. <!-- By 1984, College Board Geomarkets were incorporated as a search filter in the College Board Student Search Service \[CITE\]. --> Geomarkets also function as a student list search filter for the College Board's Student Search Service. A student list contains the contact information of prospective students who meet the search filter criteria (e.g., test score, GPA) specified by the university. Student lists are the fundamental input for undergraduate recruiting campaigns because purchased names -- alongside prospects who reach out on their own -- constitute the set of prospects who receive subsequent recruiting interventions (e.g., mail, email) designed to push them toward the application and enrollment stages of the "enrollment funnel." @RN4895 reports that 86% of public colleges and 87% of private colleges purchase student lists.[^11] Historically, dominant list vendors are the College Board and ACT, which derived student list data from test-takers, but the test-optional movement, advances in technology, and surging private equity investments have contributed to new sources of student list data [@list_biz_blinded].

@@

However, @RN5012_blinded argue that student list products may exacerbate racial inequality in college access. The authors conceptualize student list products as "selection devices" [@RN4778] that enable colleges to select which prospective students they target by incorporating search filters (e.g., high school graduating class, state). @RN4786 defines "racialized inputs" as ostensibly race-neutral inputs that are systematically correlated with race because marginalized racial/ethnic groups have historically been excluded from the input. @RN5012_blinded argue that several frequently utilized student list filters (e.g., zip code, AP test score, SAT score) meet the criteria of racialized inputs. Using a national sample of high school students and using data from actual student lists purchased by public universities, @RN5012_blinded show that racialized search filters have a strong negative relationship with the selection of Black and Latinx prospects.

Because of the extent of residential segregation in the U.S., geographic borders are a commonly studied racialized input in scholarship about algorithmic bias [@RN4775; @RN4849; @RN4887]. Student list products offer many geographic filters. Some geographic filters are based on known geographic borders (e.g., zip code, county, CBSA, state). College Board also created geographic borders that subsequently became filters in the Student Search Service, such as Geomarkets and Geodemographic Segment filters.[^14] In their sample, @RN5012_blinded found that 11% of student lists purchased by public research universities filtered on Geomarkets. Drawing from @RN4786, we argue that geomarkets satisfy the two criteria of racialized inputs. First, they are ostensibly race-neutral inputs in that neither @RN4982 -- nor subsequent EPS promotional material [e.g., @RN1621] -- mention race. Second, Geomarkets may be systematically correlated with race. For example, Geomarket "CA07" is "City of Oakland" and "CA08" is "Alameda County, excluding Oakland." <!--# A university might purchase prospect contact information by filtering on CA08 in combination with additional filters, such as year of high school graduation and SAT test score.--> We extend @RN5012_blinded by examining how the racial composition of actual student list purchases changes if universities filter on particular Geomarkets. <!--# This manuscript focuses on Geomarkets. -->

@@

The discussion of homophily by @RN4975 describes not only the logic of the Market Segment Model but also a broader institutional logic of the enrollment management industry. The logic of homophily assumes that there is a natural cultural match between students and colleges, suggesting colleges should find prospective students who resemble students who enrolled at their institution in the past and focus recruiting efforts on the same schools and communities that are popular with peer colleges. We suggest that homophily is the operating logic most consistent with college recruiting behavior observed in the wild. For example, @RN5023_blinded find great overlap in the sets of private schools visited by the University of South Carolina and the University of Alabama. <!--# The study also finds great overlap in the schools visited by Villanova University and the University of Notre Dame, with a disproportionate focus on visits to Catholic schools. When we observe that students at selective private colleges tend to be rich and that out-of-state students at public research universities tend to be rich [@RN4722], this is partly because these universities devote most of their recruiting expenditures towards students from affluent schools and communities [@RN4758; @RN5023]. --> As @RN4975 argues, observed homophily is not the result of voluntary action; rather, homophily is programmed into algorithms that create connections.

@@

Supply-side explanations of credentialism and more recent scholarship on enrollment management complement the cultural capital model of sorting students to colleges. The credentialism literature recognizes that colleges have a financial incentive for educational credentials to determine the competition for socially and economically advantageous positions [@RN865; @RN861; @RN880; @RN869] and this credentials arms race favors affluent households. Scholarship on enrollment management describes which students colleges want and how colleges go about attracting these students at different stages of the enrollment funnel [@RN3519; @RN4324; @RN3522; @RN4407; @RN4758_blinded; @RN3544; @RN3547; @cottom2017lower]. @RN3519 and @RN4321 describe a tacit arrangement between high school guidance counselors on the demand side and college admissions officers on the supply side. Counselors at well-resourced -- especially private -- schools are motivated to give their students a competitive advantage in admissions. Meanwhile, college admissions officers are motivated to enroll students who can afford full tuition price and are likely to donate in the future. These mutually beneficial desires are consummated by recognizing that upper and upper-middle class applicants satisfy the extracurricular needs of the college. The orchestra needs oboists. The lacrosse team needs players. Quantitative analyses of visits from colleges to high schools are consistent with @RN3519; selective private colleges and public research universities devote most of their recruiting resources on courting students from privileged schools and communities [@RN5023_blinded; @RN4758_blinded]. However, these analyses conceive of recruiting as something that is done solely by individual colleges.

@@

@list_biz_blinded describes four key dynamics in the enrollment management (EM) industry, with a focus on the market for student list data. First, EM consulting firms are central to the creation and implementation of recruiting campaigns [@RN5027]. The top two EM consultancies are Ruffalo Noel Levitz, which claims to serve 1,900 colleges annually [@rnlabout_2025], and EAB, which claims to serve 2,100 colleges [@about_eab]. College reliance on EM consultancies is partially explained by the high level of burnout and employee turnover in the EM profession [@hoover_em]. When we issued records requests to public universities about their student list purchases, at least 50% of universities indicated that they outsourced student list purchases to an EM consultancy. Often, the university was buying names each year but no employee had knowledge about which vendors they were buying from or which search filters they were utilizing.

@@

The third dynamic in the EM industry has been the growth of the test-optional admissions movement, which reduces both the number of paying test-takers and the coverage of student list products sold by College Board and ACT. ACT responded to uncertainty in their core testing business by attempting to become an edtech company [@blumenstyk_2019], failing [@molnar_2020a], and being acquired by Nexus Capital in 2024 [@knox_2024]. Over the last two decades, the testing companies developed new search filters that explicitly incorporated predictive analytics -- for example College Board's Geodemographic Segment filters [@RN4565] -- developed free college search engines, and entered the enrollment management consulting market more aggressively [@list_biz_blinded]. College Board has weathered the test-optional movement thanks to robust revenues from the Advanced Placement product, but the decline in PSAT/SAT/PreACT/ACT test-takers has undermined their oligopoly in the market for student list data.

The fourth -- and most important -- dynamic is the transformation of EM from an owner-operated to private-equity owned industry and the resulting eroding distinction between consultancy and software vendor. The 1980s and 1990s was a period of market entry in EM, when college admissions professionals or professors hung up a shingle [@RN5027]. The 2010s were a period of acquisitions and concentration. Drawing from resource dependence theory, @list_biz_blinded describe the proliferation of horizontal acquisitions to grow market share and vertical acquisitions designed to make customers more dependent on particular recruiting products. After regulatory scrutiny pushed private-equity interests out of the for-profit college market [@RN5022], private equity investors found value in acquiring firms that provided services to direct providers. Direct providers are subsidized by state and federal funds (Title IV financial aid) and these funds are then paid to third-party vendors. Therefore, private equity ownership of direct providers and third-party vendors are both examples of government-subsidized rent-seeking.
